---
title: "Sablefish MSE Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alaska Sablefish MSE Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A fully worked example of this MSE framework for Alaska Sablefish follows. This example includes the following features:

- multiple operating models
- multiple harvest control rules
- state-dependent and state-independent recruitment functions
- HCRs with stability constraints and harvest caps
- HCRs with different reference point levels
- Parallel MSE simulations
- MSE output processing and plotting.

### Defining Operating Models

A base operating model (OM) object, `sable_om`, is already available in the package to start from. As all of the OMs developed for this example vary only in terms of how recruitment is projected in the future, we will load this base OM object and simply modify the recruitment parameter(s).

```{r, load_base_om, eval=TRUE}
library(SablefishMSE)
data(sable_om)
```

OM1 (`om_rand_recruit`) will project future recruitment by resampling from the historical recruitment timeseries, using the `resample_recruits` function:
```{r, om1, eval=TRUE}
# Normal recruitment
data(om_rand_recruit)
om_rand_recruit$recruitment
```

OM2 (`om_bh_recruit`) projects future recruitment via a Beverton-Holt stock recruit relationship with $h=0.85$ and $\sigma_R = 1.20$. As a state-dependent recruitment function, the underlying `beverton_holt` R function is a "function factory", whose returned function is used internally by the MSE to compute recruitment based on current spawning biomass. 

```{r, om2, eval=TRUE}
data(om_bh_recruit)
om_bh_recruit$recruitment
```

OM3 (`om_bhcyclic_recruit`) projects future recruitment as occurring in distinct regimes: first a "low" regime for 20 years, followed by a "high" regime for 5 years. The regimes each have a uniquely parameterized Beverton-Holt stock recruitment relationships that differ with respect to the $R_0$ and $S_0$ parameters. This type of recruitment function is available with the provided `bevholt_regimes` function.

```{r, om3, eval=TRUE}
data(om_bhcyclic_recruit)
om_bhcyclic_recruit$recruitment
```

The four OM objects are placed in a  list that will later be passed to the MSE function.

```{r, om_list, eval=FALSE}
om_list <- afscOM::listN(om_rand_recruit, om_bh_recruit, om_bhcyclic_recruit, om_immcrash_recruit)
```

### Defining Management Procedures
Management procedures (MPs) are specified as a complex list of values that define the specifics of the harvest control rule to be applied, the reference point targets to use, ABC-TAC and TAC-landings reductions, survey frequency, and assessment frequency. A default MP object is available through the `setup_mp_options` function. The eleven MPs specified here differ only in terms of their HCR function and their reference point target, so the default MP object can, largely be reused.
```{r, setup_mp_options, eval=TRUE}
mp_base <- setup_mp_options()
mp_base
```

MP1 (`mp_f40`) is an implementation of the NPFMC Tier 3 HCR that is currently used to manage sablefish in Alaska. It is a threshold HCR that uses a target reference point of $SPR_{40\%}$. The HCR function (`tier3`) outputs in units of fishing mortality (F), as specified by "units". No harvest cap or stability constraints are applied. 
```{r, mp1, eval=TRUE}
data(mp_f40)
mp_f40
```

MP2 (`mp_f50`) is a simple variation on MP1, where the target reference points are based on $SPR_{50\%}$ instead of $SPR_{40\%}$.
```{r, mp23, eval=TRUE}
data(mp_f50)
mp_f50
```

MP3 (`mp_5perc`) is another simple variation on MP1, where a stability constraint of 5%  are applied to the ABC. Otherwise, these MPs are defined exactly as MP1. The stability constraints limit the year-to-year change in ABC to the specified percentage, such that the ABC cannot increase or decrease by more than the specified percentage from one year to the next. Stability constraints are implemented in catch units, and are thus applied after the `tier3` HCR function has output a recommended F.

MP4 (`mp_25cap`) is another simple variation on MP1, whereby a maximum permissible TAC is implemented. If the recommended TAC (via evaluation of the HCR function) exceeds the specified harvest cap, the TAC will be reduced to the harvest cap exactly. Attainment will be applied after the harvest cap has been applied. 
```{r, mp45, eval=TRUE}
data(mp_5perc)
data(mp_25cap)

mp_5perc
mp_25cap
```

The four MP objects are placed in a master list that will later be passed to the MSE function.

```{r, mp_list, eval=FALSE}
hcr_list <- afscOM::listN(
    mp_f40, mp_f50, 
    mp_5perc, mp_25cap
)
```

### Defining the `mse_options` Object
An `mse_options` list object is also defined to provide additional options to the MSE function. Critically, this list object defines: the number of years to project for, the length of the OM conditioning period, the simulation year in which to begin projecting recruitment using the provided recruitment function, and whether the estimation method should be run or not. 

The default values are all appropriate for this case study, and can be accessed using `setup_mse_options`.

```{r, mse_options, eval=TRUE}
mse_options_base <- setup_mse_options()
mse_options <- mse_options_base
mse_options$n_spinup_years <- 54
mse_options$recruitment_start_year <- 54
mse_options$n_proj_years <- 25

mse_options

mse_options_list <- afscOM::listN(mse_options)
```

### Running the MSE
Now that all of the OMs and MPs have been appropriately defined, the full set of MSE simulation can be run. Here we will run each combination of OM and MP for 20 random simulations in parallel. The parallel computing overhead is handled internally by the MSE function and uses 2 fewer than the available number of compute cores on the machine. 

```{r, run_mse, eval=FALSE}
nsims <- 20
seed_list <- sample(1:(1000*nsims), nsims)  # Draw 20 random seeds
model_runs <- run_mse_multiple(
    om_list, 
    hcr_list, 
    seed_list,
    nyears=100,
    mse_options_list=mse_options_list,
    diagnostics = TRUE,
    save=TRUE
)
```
Depending on the number of OMs, MPs, seeds, and other options, this function may take several hours to complete. A relatively fast machine, with >10 compute cores should take ~2 hours to complete all of the simulations as specified above.

### Processing MSE Results
The above MSE will generate results from 40 unique models (10 HCRs across 4 OMs), each model consisting of results for 20 simulations, with each simulation having lasted 75 years. The amount of output data often makes processing MSE results tricky. Multiple helper functions are provided by the package to make processing results easier.

Before prcoessing output, create a data.frame that specifies which models in the `model_runs` used which combination of OM and MP object. Often this can be simply accomplished using `expand.grid`, as this is what is used internally by `run_mse_multiple()`. For readability and plotting purposes, it is often useful to assign each OM and HCR a human readable name in the same order as the objects appear in their respective lists (`om_list` for OMs and `hcr_list` for MPs).
```{r, extra_columns, eval=FALSE}
# Quick Way to get the names of the OMs and HCRs in the same order as they appear in the lists
om_names <- unlist(lapply(om_list, \(x) x$name))
hcr_names <- unlist(lapply(hcr_list, \(x) x$name))

extra_columns <- expand.grid(
    om = om_names,
    hcr = hcr_names
)
```

Five helper functions are predefined to facilitate working with common outputs from the MSE: spawning biomass, fishing mortality, recruitment, ABC and TAC, and landed catch. Each function is defined very similarly, taking the list of MSE output objects (`model_runs`) and the data.frame of OMs and HCRs that apply to each model (`extra_columns`). These are used internally by each function to assign the correct OM and HCR name to each model, and pull the correct data from the MSE outputs.

```{r, data_processing, eval=FALSE}
ssb_data <- get_ssb_biomass(model_runs, extra_columns, sable_om$dem_params, hcr_filter=hcr_names, om_filter=om_names)

f_data <- get_fishing_mortalities(model_runs, extra_columns, hcr_filter=hcr_names, om_filter=om_names)

abctac <- get_management_quantities(model_runs, extra_columns, spinup_years=mse_options$n_spinup_years, hcr_filter=hcr_names, om_filter=om_names)

catch_data <- get_landed_catch(model_runs, extra_columns, hcr_filter=hcr_names, om_filter=om_names)

```

These helper functions return tibbles in long format that can be easily provided to `ggplot` for plotting or used within other tidyverse style data processing pipelines. A plotting function is also defined alongside each data processing function above, and will return a `ggplot` object that can be further modified outside of the function. For each plotting function, the parameter `v1` corresponds to the column name assigned to the "color" aesthetic, and the parameter `v2` corresponds to the column names to facet by. Facetting by multiple variables (e.g. a facet grid) is not currently supported by default.

```{r, plotting, eval=FALSE}
common_trajectory <- mse_options$n_spinup_years # the number of years of historical data before the recruitment and HCR functions begin applying

plot_ssb(ssb_data, v1="hcr", v2="om", common_trajectory=common_trajectory, show_est = FALSE)

plot_fishing_mortalities(f_data, v1="hcr", v2="om", common_trajectory = common_trajectory, show_est=FALSE)

plot_abc_tac(abctac, v1="hcr", v2="om", common_trajectory=common_trajectory)

plot_landed_catch(catch_data, v1="hcr", v2="om", common_trajectory = common_trajectory)

```

### Computing Performance Metrics
Helper functions are also provided to compute performance metrics. Individual functions for each performance metric are available (though outputs may require additional processing), or the `performance_metric_summary` function can be used to quickly compute all of them. Like with the other data processing functions, a default plotting functions for performance metrics is also available.

```{r, performance_metrics, eval=FALSE}
perf_tradeoffs <- compute_performance_metric_summary(
    model_runs, 
    extra_columns, 
    sable_om$dem_params, 
    ref_naa,
    hcr_filter=hcr_names,
    om_filter=om_names,
    interval_widths=c(0.50, 0.80),
    time_horizon = c(55, 130), 
    extra_filter = NULL,
    relative=NULL, 
    summarise_by=c("om", "hcr"),
    summary_out = FALSE,
    metric_list = c("avg_catch", "avg_variation", "avg_ssb", "avg_age", "prop_years_lowssb") 
)

perf_data <- performance_metrics$perf_data
plot_performance_metric_summary(perf_data)

```